`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-01-29 paper/paper.qmd
get rid of previous figures 1 and 2 (healthcare tt from weiss 2020 and outline of SIN). Unnecessary

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health (@hulland2019travel; @weiss2020global; Figure \ref{fig:tt_healthcare}) and a myriad others [@nelson2019suite]. Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref]. `{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest (e.g., travel time from health care facilities in Figure \ref{fig:tt_healthcare} from @weiss2020global).

@@

```{r ttexample,  include=FALSE}
library(terra)
library(malariaAtlas)

boundary_kh <- getShp(ISO = "KHM")

tt_kh <- getRaster(
  dataset_id = "Accessibility__202001_Global_Motorized_Travel_Time_to_Healthcare",
  shp = boundary_kh
)
```

```{r ttexample_image, echo=FALSE, fig.cap = "\\label{fig:tt_healthcare}Existing surface of motorised travel time from healthcare facilities, exerpted to Cambodia (Weiss et al., 2020)"}
library(ggplot2)
library(tidyterra)

ggplot() +
  geom_spatraster(
    data = tt_kh
  ) +
  scale_fill_viridis_c(
    option = "A",
    direction = -1,
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "minutes")
```

`{traveltime}` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `{traveltime}` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global as well.

@@

Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] (@fig-tt_healthcare) have generated significant interest and use (collectively >1500 citations per google scholar at the 28th of January 2025), and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

@weiss2020globa made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ). To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.

@@

```{r get basemap}
library(geodata)

@@

```{r plot basemap, fig.cap = "\\label{fig:basemap}Area of interest: Singapore."}

ggplot() +
  geom_spatvector(
    data = sin
  ) +
  scale_fill_manual(na.value = "transparent") +
  theme_minimal()
```

@@

Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin` (Figure \ref{fig:basemap}), a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.)

@@

```{r  download friction surface, echo=TRUE, output=FALSE}
library(traveltime)

friction_singapore <- get_friction_surface(
    surface = "walk2020",
    extent = sin
  )|> 
  mask(sin)
```

@@

Below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points @fig-data. Higher values of friction indicate more time travelling across a given cell.

```{r plot data, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}

library(tidyterra)
library(ggplot2)

>>>>

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and a myriad others [@nelson2019suite]. Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref]. `{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.

@@

`{traveltime}` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `{traveltime}` package, we suggest citations of the package should be accompanied by citing the underlying work [@weiss2018global; @weiss2020global] as well.

@@

Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use (collectively >1500 citations per google scholar at the 28th of January 2025), and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

@weiss2020global made their code available as an R script to allow for reproduction and extension of their analyses (<https://malariaatlas.org/wp-content/uploads/2022/11/R_generic_accessibilty_mapping_script_2020-1.txt> ). To further enable extension of this work, here we have developed an R package based on that code to seamlessly calculate the travel time from any arbitrary set of locations.

@@

```{r get basemap}
library(terra)
library(geodata)

@@

@@

Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.)

@@

```{r  download friction surface, echo=TRUE, output=FALSE}
friction_singapore <- get_friction_surface(
    surface = "walk2020",
    extent = sin
  )|> 
  mask(sin)
```

@@

Below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points (Figure \ref{fig:data}). Higher values of friction indicate more time travelling across a given cell.

```{r plot data, fig.cap = "\\label{fig:data}Friction surface raster of Singapore, showing Singapore boundary in grey, and station locations as grey points."}
library(tidyterra)
library(ggplot2)
